Long-form recording of infant body position in the home using wearable inertial sensors

### 2023-03-27 manuscript.tex
up to results

But automatic classification can only replace human annotation if it is sufficiently accurate and unbiased. An independent assessment of the LENA® algorithms found mixed results about the accuracy of different outcomes. For example, correlations between human transcribed counts of adult words and child vocalizations against LENA®'s automatic counts were strong, \emph{r} = .698 and \emph{r} = .649, respectively (Cristia et al., 2020). For other metrics, such as the number of ``conversational turns'' between the child and communicative partners and the identification of male speakers, agreement was poor. Thus, for some use cases (and for some metrics), long form audio recordings provide a mobile, unobtrusive way to automatically score real-time data over a full day. In the remainder of the paper, we turn to the question of how to replicate those key features in long form recordings of motor behavior.

@@

Accordingly, the overarching goal of the current study is to test the validity of long-form body position recording in the home during unsupervised, everyday behavior. Supervised recordings from past work (Airaksinen et al., 2022, 2020; Franchak et al., 2021), whether in the home or in the lab, let researchers set up the situation to encourage or restrict certain behaviors. Prior work focused on ``free play'', in which caregivers were asked to play with the infant without restraining the infant or otherwise shaping how they could move. However, in a real day, non-play activities (e.g., eating lunch in a high chair) create challenging situations for applying automated classification of body position. Indeed different types of sitting---independently on the floor, supported on a caregiver's lap, or restrained in a high chair---all need to be counted as equal. Moreover, models must be able to new variants of behaviors that occur over the course of natural behavior that did not appear in the training data. Will models trained on video-recorded observations at the beginning of the day generalize to predict behavior at a later time? Assessing the validity of temporally \emph{distal} periods is a crucial step to establish whether automatic classification can be used to measure body position across a day.

In the current study, we report the feasibility and validity of body position classification over the full day in the home based on 34 testing sessions from 22 infants aged 4-14 months. Participants received a custom pair of infant leggings embedded with 4 IMUs (one on each ankle and one on each hip) and a video camera to collect ground truth data about infant body position. A \emph{proximal comparison} period began when participants received the equipment and completed a guided phone call during which caregivers were asked to elicit different body positions based on prompts from the experimenter. Although not directly supervised, this period was most similar to previous recordings because it occurred during a convenient time for the infant and caregiver to play while they received instructions from the experimenter. The \textbf{first goal} of the current study was to determine the accuracy of body position classification during the proximal comparison period using this novel, semi-supervised procedure in participants' homes. Past work found better performance using ``individual models''---models that were trained on one participant's data to predict their later behavior---compared with a ``group model'' that aggregated data from all infants to create a single body position classifier (Franchak et al., 2021), so we compared both modeling approaches in the current investigation.

The crucial test was how well models predicted later behavior over longer recordings of everyday activities. A second, \emph{distal comparison} period followed the proximal comparison period and captured approximately 90 minutes of home behavior that was completely unsupervised. Caregivers and infants could (and did) do whatever they wished. Because this recording happened a considerable amount of time after the initial setup and instructions from the experimenter, accuracy could decline if caregivers or infants moved the garment or sensors. Moreover, increasing variation in everyday activities during the distal comparison creates a greater challenge, testing whether body position classification models can generalize to novel test cases. Thus, the \textbf{second goal} of our study was to assess accuracy during the distal comparison.

After the distal comparison period when video recording ceased, we asked caregivers to have infants wear the IMUs for the rest of the day until their regular bedtime, creating the \textbf{first real-time, full-day dataset of infant body position}. Interpreting such data required caregivers to log when infants napped, when they removed the sensor garment for diaper changes or other reasons, and when infants went to bed at the end of the day. Thus, the \textbf{third goal} of the study was to examine the quality of the full-day data. Could body position be successfully recorded over the desired period? The novel full-day dataset affords us a unique chance to ask whether the estimated time infants spent in different body positions align with past results using video and survey methods. Thus, we determined whether full-day body position measurements conformed to expected age differences in body position. Based on past results (Franchak, 2019), infants should spend increasingly more time sitting and upright but less time supine over the age range tested (4 to 14 months).

@@

Infants were recruited in one of two age groups: \emph{Younger} infants were between 4 and 7 months and \emph{older} infants were between 11 and 14 months. There were 8 infants in the 4-7 month group and 14 in the 11-14 month group. Ten infants were female and 12 were male. Families were recruited through social media advertisements and from community events in Southern California. Most infants were reported as either Hispanic and White or Non-Hispanic and White. Families were compensated \$30 for every home recording session they completed. The BLINDED Institutional Review Board reviewed and approved all procedures associated with the study. All caregivers gave their informed consent before the start of the study.

Most participants were tested in a single session (\emph{n} = 15), but 7 participants contributed between 2-4 sessions as part of an ongoing longitudinal study. Only 1 session was excluded due to a technical error---one of the four IMU sensors failed to record, resulting in an unusable set of data for classification. Across the two age groups, we report data on a total of 34 sessions, with 14 sessions from younger infants and 20 sessions from older infants. Across sessions, younger infants' age ranged from 3.8 to 7.2 months (\emph{M} = 5.2) and older infants' age ranged from 10.7 to 14.2 months (\emph{M} = 11.7) .

@@

Video recordings were captured using an action camera on a miniature tripod (Insta360 ONE R) that caregivers placed in the room that the infant was in. The proprietary ``Boosted Battery Base'' attached to the action camera to allow for a total of 3 hours of recording. As described below, the video recording was not continuous. The camera created two video files temporally separated by a gap of approximately 40-45 s. Caregivers also received a log sheet to record times that infants napped as well as times that the sensor garment was removed from the infant (e.g., baths, diaper changes).

@@

Figure \ref{fig:exemplar-timeline}) shows an exemplar timeline of the entire procedure and recording periods for a single participant. On the day of the visit, a researcher arrived at the participant's home in the morning and prepared all the equipment at the doorstep. In order to synchronize all three recording devices, the researcher began by first turning on the video camera and the LENA audio device (the IMUs were already configured and placed before arrival because they required a proprietary sensor dock to begin recording in the laboratory). To create an easily recognizable synchronization point between the video recording and IMU data, the researcher dropped or struck the leggings (containing the IMUs) on a surface in view of the camera, as in Franchak et al. (2021). All the equipment---once recording and with synchronization information recorded---was placed inside a large bucket and left outside the family's front door. The researcher then called the caregiver on the phone and walked them through a set of procedures needed to properly set up the equipment and record video for ground truth human annotation of body position. At the start of this ``guided call'', the caregiver was instructed to place the camera in an area that captured the majority of the room. Next, they were asked to put the pair of leggings and shirt on their infant, with the researcher providing guidance about how to correctly orient the garments.

@@

Because the camera only had the battery life to record for \textasciitilde3 hours (split into two 90-minute video files), this divided the day into different periods for analysis. As seen in the bottom of Figure \ref{fig:exemplar-timeline}), the \emph{video period} comprised the first three hours of recording starting from the researcher's arrival when they turned on the camera. The first 90-minute video file, termed the \emph{proximal comparison}, contained the activities during the guided call followed by a period of infants and caregivers resuming their normal activities. Because this video contained the synchronization point, the data in this period had high temporal synchrony between IMU and video data. Synchrony errors were less than 30-60 ms (1-2 video frames). When the first video file ended, a second video file was recorded, termed the \emph{distal comparison}. This video recorded the next 90 minutes of natural activity. However, because there was a variable gap of \textasciitilde40 s between the two videos, temporal synchrony in the distal comparison video was approximate containing offsets of \textasciitilde{} 5 s in either direction.

@@

The proximal and distal comparison videos for each participant were annotated by trained human coders to identify infant \emph{body position} into one of 5 mutually-exclusive categories following the definitions in prior work (Franchak et al., 2021): supine, prone, sitting, upright, or held by caregiver. All coding was done using Datavyu software (datavyu.org).

Supine was coded when the infant was lying on their back or was reclined up to a 45 degree angle. Supine was also coded in the rare cases when the infant was laying on their side. Prone was coded when the infant was lying on their stomach, was stationary supported by the hands/knees or the hands/feet, or was crawling. We scored sitting to include any form of the following seated positions: 1) infants sat with their buttocks on a surface, such as on the floor or a caregiver's lap, 2) infant was in a kneeling-sit position, in which their knees were on the ground with their legs tucked underneath the buttocks, and 3) infant was in a seating device, such as a high chair, that kept the torso oriented perpendicular to the ground (a reclined position, such as in a young infant's car seat, would be counted as supine). Upright was coded when the infant was standing or squatting on the ground with two feet or walking (regardless of whether infants' balance was assisted by a caregiver or with their hands holding onto something for support). When an infant was carried by a caregiver, held was coded. However, when the caregiver was sitting with the infant in their lap the infant's body position was coded as if the caregiver was a surface (e.g., if the infant was sitting on the caregiver's lap this was coded as sitting). Times during the video when the infant was out of view were excluded. Periods when the leggings were being adjusted or taken off the infant were also excluded, as were transitions between body positions.

A primary coder completed annotation for the full length of the video, while an independent reliability coder completed annotation for the first thirty minutes of each video. Interrater reliability was based on the proportion of video frames that the two coders chose the same body position code. Overall agreement averaged 90.9\% across video files, ranging from 68.4\%-100\%. Cohen's \emph{kappa} averaged 86.1\% across video files, ranging from 31.0\%-100\%.

@@

Classification training and prediction was conducted on a windowed dataset that summarized the raw, 62.5 Hz motion signals within 4-s windows. Overlapping moving windows captured 4-s (250 samples) of data starting each second, which is a common unit of analysis in prior studies of human activity classification (Airaksinen et al., 2020; Franchak et al., 2021; Nam \& Park, 2013). The 250 samples of data in each window were aggregated into a variety of motion features---summary statistics that could be fed into the machine learning model. The minimum, maximum, 25th percentile, 75th percentile, mean, median, skew, kurtosis, standard deviation, and sum were computed for each signal (e.g., right hip X linear acceleration, left ankle pitch angular acceleration). The 10 summary statistics and 24 sensor signals generated 240 columns of motion features that described movement within each window. Furthermore, a series of cross-sensor and cross-orientation summaries (such as the correlation, magnitude, and difference between pairs of sensors) added an additional 196 columns of motion features. The 436 total motion features corresponded to a single body annotation code for each 4-s window. Windows were only used for training/testing if they contained a single body position for \textgreater{} 75\% (3 s) of time within the window to ensure that motion signals could be tied to a clear example of each behavior.

The resulting windowed datasets were used for machine learning classification and validation. For each analysis reported in the results, a subset of data were defined as a ``training'' set and another, independent portion of the data were defined as a ``testing'' set. Different training/testing set combinations were used to address different goals; here we describe the general procedure of how a classifier was derived from a training set. Random forest models (Breiman, 2001) were trained to predict the human coded body position label for each window from the set of 436 motion features using the \emph{randomForest} package (Liaw et al., 2002). Random forest models use random subsets of features and random subsets of data across many iterations; this random subsetting creates models that are less likely to be overfit. The \emph{tidymodels} set of packages in R (Kuhn \& Wickham, 2020) was used to tune the ``mtry'' and ``ntrees'' hyperparameters, which control the number of columns sampled and the total number of trees created when modeling. We use an mtry parameter of 44 and and ntree parameter of 550; however, changing hyperparameters had little impact on model performance. Just like a linear regression model, the resulting random forest model could later be applied to a set of testing data with the \emph{predict} function. Validation analyses below will compare the human-coded annotations and random forest model predictions for testing data that are independent from training data, providing a test of whether models can successfully generalize.

@@

\url{https://nyu.databrary.org/volume/1580} - all raw data files and annotations for one participant

This reproducible manuscript showing how we accumulated

@@

>>>>

But automatic classification can only replace human annotation if it is sufficiently accurate and unbiased. An independent assessment of the LENA® algorithms found mixed results about the accuracy of different measurements. For example, correlations between human transcribed counts of adult words and child vocalizations against LENA®'s automatic counts were strong, \emph{r} = .698 and \emph{r} = .649, respectively (Cristia et al., 2020). For other metrics, such as the number of ``conversational turns'' between the child and communicative partners and the identification of male speakers, agreement was poor. Thus, for some use cases (and for some metrics), long form audio recordings provide a mobile, unobtrusive way to automatically score real-time data over a full day. In the remainder of the paper, we turn to the question of how to replicate those key features in long form recordings of motor behavior.

@@

Accordingly, the overarching goal of the current study is to test the validity of long-form body position recording in the home during unsupervised, everyday behavior. Supervised recordings from past work (Airaksinen et al., 2022, 2020; Franchak et al., 2021), whether in the home or in the lab, let researchers set up the situation to encourage or restrict certain behaviors. Prior work focused on ``free play'', in which caregivers were asked to play with the infant without restraining the infant or otherwise shaping how they could move. However, in a real day, non-play activities (e.g., eating lunch in a high chair) create challenging situations for applying automated classification of body position. For example different types of sitting---independently on the floor, supported on a caregiver's lap, or restrained in a high chair---all need to scored as the same behavior. Moreover, classifiers must be able to detect new variants of behaviors that might arise over the course of a real day; it is impossible for researchers to gather training data for every possible variation that might occur. Thus, our central question is whether models trained on video-recorded observations at the beginning of the day generalize to predict behavior at a later time? Assessing the validity of temporally \emph{distal} periods---when infants and caregivers are unsupervised and free to follow their everyday routines---is a crucial step to establish whether automatic classification can be used to measure body position across a day.

In the current study, we report the feasibility and validity of body position classification over the full day in the home based on 34 testing sessions from 22 infants aged 4-14 months. Participants received a custom pair of infant leggings embedded with 4 IMUs (one on each ankle and one on each hip) and a video camera to collect ground truth data about infant body position. A \emph{proximal comparison} period began when participants received the equipment and completed a guided phone call during which caregivers were asked to elicit different body positions based on prompts from the experimenter. This ``semi-supervised'' period was most similar to previous recordings because it occurred during a convenient time for the infant and caregiver to play while they received instructions from the experimenter. The \textbf{first goal} of the current study was to determine the accuracy of body position classification during the proximal comparison period using this novel, semi-supervised procedure in participants' homes. Past work found better performance using ``individual models''---models that were trained on one participant's data to predict their later behavior---compared with a ``group model'' that aggregated data from all infants to create a single body position classifier (Franchak et al., 2021), so we compared both modeling approaches in the current investigation.

The crucial test was how well models predicted later behavior over longer recordings of everyday activities. A second, \emph{distal comparison} period followed the proximal comparison period and captured approximately 90 minutes of home behavior that was completely unsupervised. Caregivers and infants could (and did) do whatever they wished, and no researcher was present. Because this recording happened a considerable amount of time after the initial setup and instructions from the experimenter, accuracy could decline if caregivers or infants moved the garment or sensors. Moreover, increasing variation in everyday activities during the distal comparison creates a greater challenge, testing whether body position classification models can generalize to novel test cases. Thus, the \textbf{second goal} of our study was to assess accuracy during the distal comparison.

After the distal comparison period when video recording ceased, we asked caregivers to have infants wear the IMUs for the rest of the day until their regular bedtime, creating the \textbf{first real-time, full-day dataset of infant body position}. Interpreting such data required caregivers to log when infants napped, when they removed the sensor garment for diaper changes or other reasons, and when infants went to bed at the end of the day. Thus, the \textbf{third goal} of the study was to examine the quality of the full-day data. Could body position be successfully recorded over the desired period? If full-day classifications of infant behavior are accurate, they should demonstrate convergent validity with other methods. Thus, we determined whether full-day body position measurements conformed to expected age differences in body position that have been identified from previous video and survey methods. Based on past results (Franchak, 2019), infants should spend increasingly more time sitting and upright but less time supine over the age range tested (4 to 14 months).

@@

Infants were recruited in one of two age groups: \emph{Younger} infants were between 4 and 7 months and \emph{older} infants were between 11 and 14 months. There were 8 infants in the 4-7 month group and 14 in the 11-14 month group. Ten infants were female and 12 were male. Families were recruited through social media advertisements and from community events in Southern California. Most infants were reported to be either Hispanic and White or Non-Hispanic and White. Families were compensated \$30 for every home recording session they completed. The BLINDED Institutional Review Board reviewed and approved all procedures associated with the study. All caregivers gave their informed consent before the start of the study.

Most participants were tested in a single session (\emph{n} = 15), but 7 participants contributed between 2-4 sessions as part of an ongoing longitudinal study. Only 1 session was excluded due to a technical error---one of the four IMU sensors failed to record, resulting in an unusable set of data for classification. Across the two age groups, we report data on a total of 34 sessions, with 14 sessions from younger infants and 20 sessions from older infants. Across sessions, younger infants' age ranged from 3.8 to 7.2 months (\emph{M} = 5.2) and older infants' age ranged from 10.7 to 14.2 months (\emph{M} = 11.7).

@@

Video recordings were captured using an action camera on a miniature tripod (Insta360 ONE R) that caregivers placed in the room that the infant was in. The proprietary ``Boosted Battery Base'' attached to the action camera to allow for a total of 3 hours of recording. As described below, the video recording was not continuous. The camera created two video files temporally separated by a gap of approximately 40-45 s. Caregivers also received a log sheet to record times that infants napped as well as times that the sensor garment was removed from the infant (e.g., baths, diaper changes, errands).

@@

Figure \ref{fig:exemplar-timeline} shows an exemplar timeline of the entire procedure and recording periods for a single participant. On the day of the visit, a researcher arrived at the participant's home in the morning and prepared all the equipment at the doorstep. In order to synchronize all three recording devices, the researcher began by first turning on the video camera and the LENA audio device (the IMUs were already configured and placed before arrival because they required a proprietary sensor dock to begin recording in the laboratory). To create an easily recognizable synchronization point between the video recording and IMU data, the researcher dropped or struck the leggings (containing the IMUs) on a surface in view of the camera, as in Franchak et al. (2021). All the equipment---once recording and with synchronization information recorded---was placed inside a large bucket and left outside the family's front door. The researcher then called the caregiver on the phone and walked them through a set of procedures needed to properly set up the equipment and record video for ground truth human annotation of body position. At the start of this ``guided call'', the caregiver was instructed to place the camera in an area that captured the majority of the room. Next, they were asked to put the pair of leggings and shirt on their infant, with the researcher providing guidance about how to correctly orient the garments.

@@

Because the camera only had the battery life to record for \textasciitilde3 hours (split into two 90-minute video files), this divided the day into different periods for analysis. As seen in the bottom of Figure \ref{fig:exemplar-timeline}), the \emph{video period} comprised the first three hours of recording starting from the researcher's arrival when they turned on the camera. The first 90-minute video file, termed the \emph{proximal comparison}, contained the activities during the guided call followed by a period during which infants and caregivers resumed their normal activities. Because this video contained the synchronization point, the data in this period had high temporal synchrony between IMU and video data. Synchronization errors were estimated to be less than 30-60 ms (1-2 video frames). When the first video file ended, a second video file was automatically recorded, termed the \emph{distal comparison}. This video recorded the next 90 minutes of natural activity. However, because there was a variable gap of \textasciitilde40 s between the two videos, synchronization in the distal comparison video was coarser, with estimated temporal offsets of \textasciitilde5 s in either direction.

@@

The proximal and distal comparison videos for each participant were annotated by trained human coders to identify infant \emph{body position} into one of 5 mutually-exclusive categories following the definitions in prior work (Franchak et al., 2021): supine, prone, sitting, upright, or held by caregiver. All coding was done using Datavyu software (datavyu.org). A Databrary repository contains the entire video recording, coding files, and raw IMU data for a single participant (\url{https://nyu.databrary.org/volume/1580}).

Supine was coded when the infant was lying on their back or was reclined up to a 45 degree angle. Supine was also coded in the rare cases when the infant was laying on their side. Prone was coded when the infant was lying on their stomach, was stationary supported by the hands/knees or the hands/feet, or was crawling. We scored sitting to include any form of the following seated positions: 1) infants sat with their buttocks on a surface, such as on the floor or a caregiver's lap, 2) infant was in a kneeling-sit position, in which their knees were on the ground with their legs tucked underneath the buttocks, and 3) infant was in a seating device, such as a high chair, that kept the torso oriented perpendicular to the ground (a reclined position, such as in a young infant's car seat, would be counted as supine). Upright was coded when the infant was standing or squatting on the ground with two feet or walking (regardless of whether infants' balance was assisted by a caregiver or with their hands holding onto something for support). When an infant was carried by a caregiver, held was coded. However, when the caregiver was sitting with the infant in their lap the infant's body position was coded as if the caregiver was a surface (e.g., if the infant was sitting on the caregiver's lap this was coded as sitting). Times during the video when the infant was out of view were excluded. Periods when the sensor garment was adjusted or taken off the infant were also excluded, as were transitions between body positions.

A primary coder completed annotation for the full length of the video, while an independent reliability coder completed annotation for the first thirty minutes of each video. Interrater reliability was based on the proportion of video frames that the two coders chose the same body position code. Overall agreement averaged 90.9\% across video files, ranging from 68.4\%-100\% for individual video files. Cohen's kappa averaged 86.1\% across video files, ranging from 31.0\%-100\% for individual video files.

@@

Classification training and prediction was conducted on a windowed dataset that summarized the raw, 62.5 Hz motion signals within 4-s windows. Overlapping moving windows captured 4-s (250 samples) of data starting each second, which is a common unit of analysis in prior studies of human activity classification (Airaksinen et al., 2020; Franchak et al., 2021; Nam \& Park, 2013). The 250 samples of data in each window were aggregated into a variety of motion features---summary statistics that could be fed into the machine learning model. The minimum, maximum, 25th percentile, 75th percentile, mean, median, skew, kurtosis, standard deviation, and sum were computed for each signal (e.g., right hip X linear acceleration, left ankle pitch angular acceleration). The 10 summary statistics and 24 sensor signals generated 240 columns of motion features that described movement within each window. Furthermore, a series of cross-sensor and cross-orientation summaries (such as the correlation, magnitude, and difference between pairs of sensors) added an additional 196 columns of motion features. The 436 total motion features corresponded to a single body annotation code for each 4-s window. Windows were only used for training/testing if they contained a single body position for \textgreater{} 75\% (3 s) of time within the window to ensure that motion signals could be linked to an unambiguous example of each behavior.

The resulting windowed dataset was used for machine learning classification and validation. For each analysis reported in the results, a subset of data were defined as a ``training'' set and another, independent portion of the data were defined as a ``testing'' set. Different training/testing set combinations were used to address different goals; here we describe the general procedure of how a classifier was derived from a training set. Random forest models (Breiman, 2001) were trained to predict the human coded body position label for each window from the set of 436 motion features using the \emph{randomForest} package (Liaw et al., 2002). Random forest models use random subsets of features and random subsets of data across many iterations; this random subsetting creates models that are less likely to be overfit. The \emph{tidymodels} set of packages in R (Kuhn \& Wickham, 2020) was used to tune the ``mtry'' and ``ntrees'' hyperparameters, which control the number of columns sampled and the total number of trees created when modeling. We use an mtry parameter of 44 and and ntree parameter of 550; however, changing hyperparameters had little impact on model performance. Just like a linear regression model, the resulting random forest model could later be applied to a set of testing data with the \emph{predict} function. Validation analyses below will compare the human-coded annotations and random forest model predictions for testing data that are independent from training data, providing a test of whether models can successfully generalize.

@@

Three online repositories contain openly shared data, materials, and analysis code. A Databrary repository (\url{https://nyu.databrary.org/volume/1580}) includes an exemplar participant's recording session, with the raw video data files, the Datavyu annotations of those video files, a log file with machine-readable synchronization points and nap/diaper change times, and accelerometer and gyroscope data for each of the 4 sensors. A CodeOcean capsule (LINK) hosts an R environment with the exemplar participant's data and source code to: 1) synchronize IMU and video annotations, 2) calculate windowed motion features for their data, and 3) train and test the body position classifier using an ``individual model''. The chosen participant's classification accuracy was at the median level. Because of the overall size of the full data set and computational power/time required to synchronize and create windowed datasets, it would not be feasible to create a reproducible environment to process the entire dataset. However, in a second CodeOcean capsule (link) we share the entire dataset of windowed motion features with corresponding body position codes used to validate the method (described in the results below), along with the processed datafiles used for each set of analyses. This reproducible manuscript created in \emph{papaja} (Aust \& Barth, 2022) can be executed in the CodeOcean environment for full computational transparency.

@@

\leavevmode\vadjust pre{\hypertarget{ref-papaja}{}}%
Aust, F., \& Barth, M. (2022). \emph{{papaja: Prepare reproducible APA journal articles with R Markdown}}. \url{https://github.com/crsh/papaja}
